---
title: "Fluctuation Pattern over the map"
author: "Joonsuk Kang"
date: "2020-05-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Here we plot the first 30 fluctuation patterns, learned using observation and simulation data, over the world map. 

For a given fluctuation pattern, we have values corresponding to the 900 predictors, each of which is the standardized tempreature from (longitude, latitude, month) where longitude $\in [85,275]$, latitude $\in [-55,65]$, and month $\in \{July, August, September, October\}$.

For each factor, we first show one figure for month-averaged patterns and then four month-specific patterns. In general, the four month-specific figures are consistent with each other, thus also with the month-averaged figure.

This temporal consistency is surprising because when identifying patterns (PCA) we did not encode any information about the four predictors corresponding to the same geographical location. Also surprising is the smoothness over locations: we did not utilize any spatial structure in PCA, either.

Figures are presented in two methods: one set of figuers in tile plots and the other in dots. The values for a fluctuation pattern are not sparse, unlike the estimated coefficients from lasso. Thus, to hightlight the continuous variation across locations, figures are shown in tile versions in addition to point plots.



## Data

```{r}
library(tidyverse)
library(glmnet); library(doMC); registerDoMC(cores=6)

# precipitation data from abby
X.o <- as.matrix(read_csv("data/forecasting_data/data/X_obs.csv"))
X.s <- as.matrix(read_csv("data/forecasting_data/data/X_sim.csv"))
Y.o <- as.matrix(read_csv("data/forecasting_data/data/y_obs.csv"))
Y.s <- as.matrix(read_csv("data/forecasting_data/data/y_sim.csv"))

# index 
whereisit <- read_csv("data/maps_data/sst_loc.csv")
whereisit %>% mutate(month=factor(month),) -> whereisit

# test set
X.test <- X.o[52:79,] # years 1991-2018
Y.test <- Y.o[52:79]
 
# train set with only observation data
X.train0 <- X.o[1:51,] # years 1940-1990
Y.train0 <- Y.o[1:51]

Y.s.array <- array(0, dim=c(51,1,40)) # use only first 51 rows: corresponding to the same years 1940-1990
X.s.array <- array(0, dim=c(51,900,40))
for (i in 1:40){
  Y.s.array[,,i] <- Y.s[Y.s[,2]==i,1][1:51]
  X.s.array[,,i] <- X.s[X.s[,901]==i,1:900][1:51,]
}
for (i in 1:40){ # standardize (mean 0, sd 1)
  Y.s.array[,,i] <- scale(Y.s.array[,,i])
  for (j in 1:900){ # standardize each column for each simulation
    X.s.array[,j,i] <- scale(X.s.array[,j,i])
  }
}

# train set for obs+simulation data
X.train.p <- X.train0 
Y.train.p <- Y.train0
for (i in 1:40){
  X.train.p <- rbind(X.train.p, X.s.array[,,i])
  Y.train.p <- c(Y.train.p, Y.s.array[,,i])
}

# SVD
svd.p <- svd(X.train.p) # obs+sim train set
mat.F.all <- t(svd.p[['v']])

```



## Figures: tiles


```{r figtiles}
library(maps); library(mapproj); library(RColorBrewer)
map <- map_data('world2')

for (factor.idx in 1:30){

pattern <- mat.F.all[factor.idx,]
df.fig <- cbind(whereisit, data.frame(pattern=pattern))

# averaged over month
df.fig %>% group_by(lat, lon) %>%
  summarise(pattern=mean(pattern)) %>%
  mutate(size=abs(pattern),
         sign=factor(sign(pattern), levels=c(1,-1), labels=c('positive', 'negative')))  %>%
  ggplot()+
  geom_tile(aes(x=lon, y=lat, fill=pattern))+
  scale_fill_distiller(palette="RdBu")+
  geom_polygon(aes(x=long, y=lat, group = group), col='black', fill=NA, data=map) +
  coord_map(xlim = c(80, 280),ylim = c(-60, 70))+
    ggtitle(paste0("Fluctuation Pattern", " #", factor.idx, ": averaged over Aug-Sept"))-> temp.fig

print(temp.fig)

# by month

df.fig %>% 
  mutate(size=abs(pattern),
         sign=factor(sign(pattern), levels=c(1,-1), labels=c('positive', 'negative')))  %>%
  ggplot()+
    geom_tile(aes(x=lon, y=lat, fill=pattern))+
    scale_fill_distiller(palette="RdBu")+
    geom_polygon(aes(x=long, y=lat, group = group), col='black', fill=NA, data=map) +
    scale_size_continuous(range=c(0,1))+
      facet_grid(~month)+coord_map(xlim = c(80, 280),ylim = c(-60, 70))+
    ggtitle(paste0("Fluctuation Pattern", " #", factor.idx, ": by month")) -> temp.fig

print(temp.fig)

}
  
```



## Figures: points

```{r figpoints}
for (factor.idx in 1:30){

pattern <- mat.F.all[factor.idx,]
df.fig <- cbind(whereisit, data.frame(pattern=pattern))

# averaged over month
df.fig %>% group_by(lat, lon) %>%
  summarise(pattern=mean(pattern)) %>%
  mutate(size=abs(pattern),
         sign=factor(sign(pattern), levels=c(1,-1), labels=c('positive', 'negative')))  %>%
  ggplot()+
  geom_point(aes(x=lon, y=lat, size=size, col=sign))+
  geom_polygon(aes(x=long, y=lat, group = group), col='black', fill=NA, data=map) +
  scale_size_continuous(range=c(0,1))+coord_map(xlim = c(80, 280),ylim = c(-60, 70))+
    ggtitle(paste0("Fluctuation Pattern", " #", factor.idx, ": averaged over Aug-Sept")) -> temp.fig
  
print(temp.fig)

# by month
df.fig %>% 
  mutate(size=abs(pattern),
         sign=factor(sign(pattern), levels=c(1,-1), labels=c('positive', 'negative')))  %>%
  ggplot()+
    geom_point(aes(x=lon, y=lat, size=size, col=sign))+
    geom_polygon(aes(x=long, y=lat, group = group), col='black', fill=NA, data=map) +
    scale_size_continuous(range=c(0,1))+
    facet_grid(~month)+coord_map(xlim = c(80, 280),ylim = c(-60, 70))+
    ggtitle(paste0("Fluctuation Pattern", " #", factor.idx, ": by month")) -> temp.fig

print(temp.fig)

}
  
```
